\section{Introduction}
\label{sec:intro}

LLM agents have been shown to discover new mathematical
constructions and algorithmic
ideas~\cite{fawzi2024funsearch,deepmind2025alphaevolve}. On
continuous constrained-optimization problems, however, it is
unclear whether an LLM-generated schedule that outperforms a
baseline contributes (a) a genuinely novel \emph{parameterization}
of the search space or (b) a particular point inside a rich family
that non-LLM search could have reached at matched compute.
Same-distribution held-out benchmarks~\cite{yao2025heurigym} cannot
distinguish these: when an LLM's score is reproduced by random
search or differential evolution \emph{inside the same
parameterization}, the LLM's contribution is the parameterization,
not the score. We propose a benchmark that isolates this
distinction: physics-based held-out evaluation and matched-compute
non-LLM ablations within each LLM-proposed family.

We present FunWake, a benchmark with three properties:

\begin{enumerate}[nosep]
    \item \textbf{Minimal interface.} The LLM writes a single function
    with 4 inputs and 4 outputs:
    \texttt{schedule\_fn(step, total\_steps, lr0, alpha0)} $\to$
    \texttt{(lr, alpha, beta1, beta2)}. A fixed skeleton handles
    initialization, gradients, and Adam updates. Each evaluation takes
    30--60 seconds on CPU.
    \item \textbf{Physics-based held-out evaluation.} The held-out
    farm differs from training in turbine model, polygon geometry,
    turbine count, and wind resource. The LLM never sees held-out AEP.
    Additionally, the training farm is evaluated at seven turbine
    counts ($N=30$--$100$) to test scaling generalization.
    \item \textbf{Low cost.} Evaluating a submitted schedule is free
    (CPU only). Running an LLM agent through the full benchmark costs
    \$5--50 depending on model. Full paper reproduction costs
    $\sim$\$500.
\end{enumerate}

\paragraph{The strongest agents, and the optimum they share.}
Restricted to feasible held-out layouts, Gemini CLI reaches
4272.7\,GWh (+26.0 over the 500-start SGD baseline of 4246.7) and
Claude Code reaches 4271.5\,GWh (+24.8). Four non-LLM baselines at
matched compute---random search over a parameterized schedule
family (4268.6), bump-family DE seeded with Claude (4272.9),
bump-family DE seeded with zero bumps (4271.9), and CMA-ES
warm-started at Claude's layout (4272.2)---reach the same 4.3\,GWh
band. A substrate-independent CMA-ES cold-started from a generic
feasible init reaches only 4189.3\,GWh, $\sim$80\,GWh below the
cluster at a matched 4500-evaluation budget.
The picture that emerges (Sec.~\ref{sec:ablations}) is a
\emph{shared local optimum} at 4272.7\,GWh: substrate-independent
CMA-ES confirms this is a real landscape feature (warm-start:
$+0.7$ in 6000 evals) rather than a shared-substrate artifact,
though we cannot rule out higher optima reachable by longer
substrate-independent runs. Several schedules produce higher
\emph{nominal} held-out AEP but violate constraints;
feasibility-aware evaluation is essential.

\paragraph{What the LLMs contribute: parameterization, not score.}
The two frontier agents propose qualitatively different
parameterizations of the schedule space: Claude Code's
14-parameter dual-Gaussian-bump family (two bump amplitudes,
centers, widths, LR, warmup, penalty coupling, quadratic boost,
two Adam betas) and Gemini CLI's sinusoidal-shake schedule plus a
separate 6-parameter 2D lattice family. Running matched-compute
differential evolution \emph{inside} Claude's bump family reaches
4271.9\,GWh starting from a zero-bump baseline seed (+24.7 over
baseline), within seed noise of the LLM-deployed schedule itself.
The LLM's contribution is the parameterization---rich enough to
contain the optimum---and not the particular point inside it.

The specific mechanisms the LLMs assemble---dual Gaussian LR bumps
coupled to an in-phase penalty dip (Claude), or a high-frequency
decaying sinusoidal perturbation paired with an aggressive
quadratic penalty ramp (Gemini)---are novel relative to standard
schedule building blocks (cosine annealing with warm
restarts~\cite{loshchilov2017sgdr}, cyclical learning
rates~\cite{smith2017cyclical,smith2019superconvergence}, noise
injection~\cite{jastrzebski2018factors}, weight
averaging~\cite{izmailov2018swa}). A controlled ablation varying penalty weight while holding the
LR schedule fixed shows that the penalty trajectory affects
held-out \emph{quality} (11.2\,GWh span) but not feasibility---the
LR dynamics alone produce feasible layouts at all penalty scales
(Sec.~\ref{sec:discussion}).

\paragraph{Contributions.}
(1)~An open-source, low-cost benchmark for LLM-discovered optimizer
schedules with physics-based held-out evaluation, matched-compute
non-LLM ablations, and scalable difficulty.
(2)~Empirical isolation of \emph{parameterization discovery} as the
load-bearing LLM contribution: matched-compute DE inside the LLM's
proposed families reaches the same optimum as the LLM-deployed
schedules.
(3)~Documentation of the LLM-proposed schedule structures (Gaussian
bumps, decaying sinusoidal shake, 6-parameter 2D lattice) together
with positive transfer from DEI to a single held-out polygon (ROWP)
across turbine counts $N=30$--$80$.
