\section{Related Work}
\label{sec:related}

\paragraph{LLMs as algorithm designers.}
FunSearch~\cite{fawzi2024funsearch} introduced the paradigm of using LLMs to evolve programs, discovering new mathematical constructions for the cap set and online bin-packing problems.
AlphaEvolve~\cite{deepmind2025alphaevolve} scaled this approach with Gemini, improving algorithms in matrix multiplication, scheduling, and geometry.
ReEvo~\cite{ye2024reevo} applied LLMs as hyper-heuristics for combinatorial optimization (TSP, CVRP, bin packing), generating operator heuristics that outperform hand-designed alternatives.
EoH (Evolution of Heuristics)~\cite{liu2024eoh} similarly evolves heuristic functions using LLM-generated mutations.
These works demonstrate that LLMs can produce functional optimization code, but evaluate generated programs on instances drawn from the same distribution used during search---a limitation when the goal is to assess algorithmic generalization rather than instance-specific tuning.

\paragraph{Benchmarks for LLM optimization agents.}
HeuriGym~\cite{yao2025heurigym} provides a benchmark suite of combinatorial optimization problems for evaluating LLM-generated heuristics, spanning graph problems, scheduling, and packing.
CO-Bench~\cite{cobench2025} standardizes evaluation across combinatorial optimization tasks with controlled difficulty levels.
LLaMEA~\cite{llamea2024} proposes an evolutionary algorithm framework where LLMs generate and mutate algorithm components.
These benchmarks focus on discrete combinatorial problems.
None evaluate LLM-generated code on continuous constrained engineering optimization problems with held-out instances that differ in physical parameters (turbine specifications, domain geometry, environmental conditions).

\paragraph{Wind farm layout optimization.}
The wind farm layout optimization problem (WFLOP) places $N$ turbines inside a polygon to maximize annual energy production subject to minimum spacing and boundary constraints~\cite{mosetti1994optimization,grady2005placement}.
The problem is non-convex with many local optima due to complex wake interactions.
Gradient-based methods~\cite{thomas2023improving,quick2024design} exploit automatic differentiation through analytical wake models but are sensitive to initialization.
Evolutionary approaches~\cite{chen2013wind,rios2023population} handle the multimodal landscape through population diversity but scale poorly.
Hybrid methods combining gradient refinement with population-based search~\cite{stanley2019massive} represent the current state of the art for large farms.
TopFarm~\cite{pedersen2019topfarm} provides an open-source framework with a constrained SGD solver that uses ADAM momentum and adaptive penalty ramping for boundary and spacing constraints.
The SGD approach with differentiable wake models has been shown to be competitive with evolutionary methods at significantly lower computational cost~\cite{quick2024design}.

\paragraph{The gap.}
No prior work evaluates LLM-generated optimization code on engineering problems with a rigorous held-out test set where physical parameters differ from training.
FunSearch and AlphaEvolve evaluate on the same problem class; ReEvo and HeuriGym use same-distribution test instances.
For wind farm optimization specifically, there is no prior work on LLM-generated layout optimizers.
Our framework is, to our knowledge, the first to evaluate whether LLM agents discover generalizable optimization algorithms versus overfit to a specific problem instance.
