\documentclass{article}

% ICML 2026 AI Scientists Workshop (4-8 pages)
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}

\title{FunWake: A Low-Cost Benchmark for LLM-Discovered\\Optimizer Schedules}

\author{
  Julian Quick \\
  DTU Wind and Energy Systems \\
  Technical University of Denmark \\
  \texttt{juqu@dtu.dk}
}

\begin{document}
\maketitle

\begin{abstract}
We introduce FunWake, a benchmark for evaluating whether LLM agents can
discover novel optimization schedules. The task is minimal: write a
Python function that returns four scalars (learning rate, penalty weight,
and two Adam momentum parameters) at each step of a fixed gradient
optimizer for wind farm layout. Each evaluation takes 30--60 seconds on
a CPU with no GPU required. A physics-based held-out farm---with
different turbine model, polygon geometry, and wind resource---catches
overfitting that same-distribution evaluation misses.

We evaluate four LLM agents (Claude Code, Gemini~2.5 Flash via
Gemini~CLI, Qwen~2.5 Coder~32B, Llama~3.3~70B) on the benchmark. When
restricted to schedules that produce \emph{feasible} held-out layouts,
the best Gemini and Claude runs are essentially tied at $\sim$4272\,GWh
(+28\,GWh over a 500-multi-start SGD baseline of 4243.75\,GWh). Claude
Code's dual Gaussian learning rate bumps with coordinated penalty
relaxation (+27.8\,GWh feasible ROWP) and Gemini's full-optimizer
hexagonal-init wrappers (+28.9\,GWh) reach comparable held-out scores
through qualitatively different mechanisms. Several schedules with
higher \emph{nominal} ROWP AEP produce infeasible held-out layouts,
underscoring the importance of feasibility checking. Across DEI farm variants with
$N=30$--$80$ turbines, the LLM-discovered schedules track or exceed the
500-multi-start baseline at every scale. FunWake is open-source, costs
\$5--50 per model to run, and evaluates submitted schedules for free
on CPU.
\end{abstract}

\input{sections/introduction}
\input{sections/related_work}
\input{sections/methods}
\input{sections/results}
\input{sections/discussion}

\section{Conclusion}
\label{sec:conclusion}

FunWake is a low-cost, open-source benchmark for LLM-discovered
optimizer schedules. The task is simple---write a 4-output Python
function---and evaluation is fast (30--60~seconds on CPU). Two findings
justify the benchmark's design: (1)~LLM agents discover genuinely novel
schedule structures (decaying sinusoidal perturbation, dual Gaussian
bumps) that generalize across farm sizes the agent never trained on, and
(2)~physics-based held-out evaluation is essential, as training AEP
improvements of +9~GWh can correspond to held-out degradation.

The benchmark is not saturated: the gap between the best schedule
(Gemini, +81.3~GWh on held-out) and the worst successful schedule
(Qwen, +4.7~GWh) spans an order of magnitude, and the discovered
schedules differ qualitatively. We release all 700+ generated schedules,
the evaluation harness, and seven farm variants at
\url{https://github.com/kilojoules/funwake}.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
